Compositional equivalences based on Open pNets

### 2021-09-21 2016-Weak-FH-Bisim/WeakBisim.tex
Update WeakBisim.tex
seance avec Rabea...: british, state vars, etc

\documentclass{lmcs}
%\overfullrule=1mm
\usepackage[multiple]{footmisc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{DotArrow}
\usepackage{hyperref}
\usepackage{listings}
\lstset{basicstyle=\small}

@@

In the nineties, several 
works extended the basic behavioural models based on labelled
transition systems to address value-passing or parameterised systems, using
various symbolic encodings of the
transitions~\cite{deSimone85,Larsen87}. 
These works use the term \emph{parameter} to designate systems where variables defining the system have a strong influence on their behaviour: in parametrised systems, parameters can typically influence the number of processes in the system or the way they interact.
In \cite{IngolfsdottirL:2001,HennessyLin:TCS95}, Lin, Ingolfsdottir and Hennessy developed a full hierarchy of
bisimulation equivalences, together with a proof system, for value passing CCS,
including notions of symbolic behavioural semantics and various symbolic 
bisimulations (early and late, strong and weak, and their congruent versions). They also extended this work to models with explicit assignments \cite{Linconcur96}.
Separately J. Rathke~\cite{HennessyRathke:TCS98} defined another symbolic semantics for 
a parameterised broadcast calculus, together with strong and weak bisimulation
equivalences, and developed a symbolic model-checker based on a tableau
method for these processes. 30 years later, no
practical verification approach and no verification platform are
using this kind of approaches to provide proof methods for
value-passing processes or open process expressions.

@@

\subsubsection*{pNets and open automata}
pNets is a convenient model to model concurrent systems in a hierarchical and parametrised way. The coordination between processes is expressed as synchronisation vectors that allow the definition of complex and expressive synchronisation patterns.
Open pNets are pNets for which some elements in the hierarchy are still undefined, such undefined elements are called \emph{holes}; a hole can be \emph{filled} later by providing another pNet.
 The semantics of pNets can be expressed as a translation to a labelled transition system, but only if the pNet has no parameter and no holes. Adding parameters to a LTS is quite standard but enabling holes inside LTSs is not a well-defined notion. 
We thus define \emph{open automata} that can be seen as LTSs with parameters and holes. The transitions of open automata are much more complex than transitions of an LTS as the firing on a transition depends on parameters and actions that are symbolic. This article defines the notion of \emph{open transition} a transition that is symbolic in terms of parameters and coordinated actions.

@@

One of the original aspects of FH-bisimulation is due to the symbolic nature of open automata. Indeed, a single state of the automaton represents a potentially infinite number of concrete states, depending on the value of state variables, and a single open transition of the automaton may also be instantiated with an unbounded number of values for the transition parameters. Consequently it would be too restrictive to impose that each transition of one automaton is matched by exactly one transition of the bisimilar automaton. Thus the definition of bisimulation requires that, for each open transition of one automaton, there exists a matching  set of open transitions covering the original one, indeed depending on the value of action parameters or state variables different open transitions might simulate the original one provided the reached states are bisimilar.

The second original aspect is the nature of the bisimulation relation.
 A classical relation between states can be seen as a function mapping pairs of state to a boolean value (true if the states are related, false if they are not). A FH-bisimulation relation maps pairs of states to boolean expressions that use variables of the two states.  Each variable used in $\Pred_{s,t}$ should refer to a state variable of $s$ or $t$ and there must be no ambiguity on the referenced variable, we thus require that the two compared automata have disjoint state variables.
Formally, a relation over the states of two open automata  $\mylangle J,\mathcal{S}_1, s_0,V_1,
   \mathcal{T}_1 \myrangle$ and $\mylangle J,\mathcal{S}_2,t_0,V_2, \mathcal{T}_2 \myrangle$ has the signature $\mathcal{S}_1\times\mathcal{S}_2\to\AlgB$. 
We adopt a notation similar to standard relations and denote it
 $\mathcal{R}=\{(s,t|\Pred_{s,t})\}$. 
The relation $\mathcal{R}$ must verify the following constraints: 1) For any pair $(s,t)\in \mathcal{S}_1\times \mathcal{S}_2$, there is a 
   single
      $(s,t|\Pred_{s,t})\in\mathcal{R}$  stating that $s$ and $t$ are related 
      if $\Pred_{s,t}$       is 
      True, i.e. the states are related when the value of the variables in $s$ and $t$  verify the predicate $\Pred_{s,t}$. 2) The free variables of  $\Pred_{s,t}$ belong to $V_1$ and $V_2$, i.e. $\vars(\Pred_{s,t})\subseteq V_1\cup V_2$, and the variables of the two open automata are disjoint, i.e.  $V_1\cap V_2=\emptyset$.
 FH-bisimulation is defined formally: 
 \begin{definition}[Strong FH-bisimulation]\label{def-FH-bisim} ~\\
\noindent
\begin{minipage}{0.67\linewidth} 	Suppose 
   $A_1 = \mylangle J,\mathcal{S}_1, s_0,V_1,
   \mathcal{T}_1 \myrangle$ and $A_2 = \mylangle J,\mathcal{S}_2,t_0,V_2, \mathcal{T}_2 \myrangle$
   are open automata with identical holes of the same sort, with disjoint sets of variables.

@@

The definition of bisimulation for open automata allows us to derive the characterisation and properties of a
bisimulation relation for open pNets. As pNets are composition
structures, it then makes sense to prove composition lemmas: we prove
that the composition of strongly bisimilar pNets are themselves
bisimilar.

@@

The initial state is defined as:
$InitState(\pNet) = \triangleleft {{s_i}_0}^{i\in L}  \triangleright$.
\end{definition}
To be precise, the state of each pLTS is entirely characterised by both the state of the automaton, and the value of its variables $V$. Consequently, the complete characterization of the state of a pNet should take into account the value of its variables $vars(P)$.

@@

Methodologies for reasoning about abstract semantics of open systems can be found in \cite{BaldanBB:2002, BaldanBB:07,Dubut:20}, authors introduce  behavioural equivalences for open systems from various symbolic approaches. Working in the setting of process calculi, some close relations  exist with the work  of the authors of \cite{BaldanBB:2002,BaldanBB:07}, where both approaches are based
on some kinds of labelled transition systems. The distinguishing feature of their approach is the transitions systems are labelled with logical formulae that provides  an abstract characterisation of the structure that a hole must possess and of the actions it can perform in order to allow a transition to fire. Logical formulae are suitable formats that capture the general class of components that can act as the placeholders of the system during its evolution. In our approach we purposely leave the algebra of action terms undefined but the only operation we allow on action of holes is the comparison with other actions. Defining properly the interaction between a logical formulae in the action and the logics of the pNet composition seems very difficult.

@@

\noindent ($\Leftarrow$) We proceed by structural induction on the rules  building  the weak transition (as described in the original definition). The recurrence hypothesis being that the  the original definition implies the characterisation (1),with the conditions stated at the bottom of the theorem. We consider the different rules:

@@

With the above lemma, we can use the same technique as for Theorem \ref{thm-equiv}  to prove that a weak FH-bisimulation is an equivalence. Indeed, we essentially use the same proof-scheme the main difference concerns  $\beta$ and  $\gamma$. Indeed, while the schema of the proof of transitivity was not directly applicable on the definition of weak bisimulation,  Lemma~\ref{lem-alternative-weak-bisim} provides a characterisation of weak bisimulation similar to the definition of strong bisimulation, and thus the same proof scheme is directly applicable.

>>>>

\documentclass{lmcs}
%\overfullrule=1mm
\usepackage[multiple]{footmisc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{geometry}                % See geometry.pdf to learn the layout
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{DotArrow}
\usepackage{hyperref}
\usepackage{listings}
\lstset{basicstyle=\small}

@@

In the nineties, several 
works extended the basic behavioural models based on labelled
transition systems to address value-passing or parameterised systems, using
various symbolic encodings of the
transitions~\cite{deSimone85,Larsen87}. 
These works use the term \emph{parameter} to designate systems where variables defining the system have a strong influence on their behaviour: in parameterised systems, parameters can typically influence the number of processes in the system or the way they interact.
In \cite{IngolfsdottirL:2001,HennessyLin:TCS95}, Lin, Ingolfsdottir and Hennessy developed a full hierarchy of
bisimulation equivalences, together with a proof system, for value passing CCS,
including notions of symbolic behavioural semantics and various symbolic 
bisimulations (early and late, strong and weak, and their congruent versions). They also extended this work to models with explicit assignments \cite{Linconcur96}.
Separately J. Rathke~\cite{HennessyRathke:TCS98} defined another symbolic semantics for 
a parameterised broadcast calculus, together with strong and weak bisimulation
equivalences, and developed a symbolic model-checker based on a tableau
method for these processes. 30 years later, no
practical verification approach and no verification platform are
using this kind of approaches to provide proof methods for
value-passing processes or open process expressions.

@@

\subsubsection*{pNets and open automata}
pNets is a convenient model to model concurrent systems in a hierarchical and parameterised way. The coordination between processes is expressed as synchronisation vectors that allow the definition of complex and expressive synchronisation patterns.
Open pNets are pNets for which some elements in the hierarchy are still undefined, such undefined elements are called \emph{holes}; a hole can be \emph{filled} later by providing another pNet.
 The semantics of pNets can be expressed as a translation to a labelled transition system, but only if the pNet has no parameter and no holes. Adding parameters to a LTS is quite standard but enabling holes inside LTSs is not a well-defined notion. 
We thus define \emph{open automata} that can be seen as LTSs with parameters and holes. The transitions of open automata are much more complex than transitions of an LTS as the firing on a transition depends on parameters and actions that are symbolic. This article defines the notion of \emph{open transition} a transition that is symbolic in terms of parameters and coordinated actions.

@@

One of the original aspects of FH-bisimulation is due to the symbolic nature of open automata. Indeed, a single state of the automaton represents a potentially infinite number of concrete states, depending on the value of the automaton variables, and a single open transition of the automaton may also be instantiated with an unbounded number of values for the transition parameters. Consequently it would be too restrictive to impose that each transition of one automaton is matched by exactly one transition of the bisimilar automaton. Thus the definition of bisimulation requires that, for each open transition of one automaton, there exists a matching  set of open transitions covering the original one, indeed depending on the value of action parameters or automaton variables different open transitions might simulate the original one provided the reached states are bisimilar.

The second original aspect is the nature of the bisimulation relation.
 A classical relation between states can be seen as a function mapping pairs of state to a boolean value (true if the states are related, false if they are not). A FH-bisimulation relation maps pairs of states to boolean expressions that use variables of the two states.  %Each variable used in $\Pred_{s,t}$ should refer to a state variable of $s$ or $t$ and there must be no ambiguity on the referenced variable, we thus require that the two compared automata have disjoint state variables.
Formally, a relation over the states of two open automata  $\mylangle J,\mathcal{S}_1, s_0,V_1,
   \mathcal{T}_1 \myrangle$ and $\mylangle J,\mathcal{S}_2,t_0,V_2, \mathcal{T}_2 \myrangle$ has the signature $\mathcal{S}_1\times\mathcal{S}_2\to\AlgB$. 
We adopt a notation similar to standard relations and denote it
 $\mathcal{R}=\{(s,t|\Pred_{s,t})\}$, where: 1) For any pair $(s,t)\in \mathcal{S}_1\times \mathcal{S}_2$, there is a 
   single
      $(s,t|\Pred_{s,t})\in\mathcal{R}$  stating that $s$ and $t$ are related 
      if $\Pred_{s,t}$       is 
      True, i.e. the states are related when the value of the automata variables% in $s$ and $t$
  verify the predicate $\Pred_{s,t}$. 2) The free variables of  $\Pred_{s,t}$ belong to $V_1$ and $V_2$, i.e. $\vars(\Pred_{s,t})\subseteq V_1\cup V_2$, and the variables of the two open automata are disjoint, i.e.  $V_1\cap V_2=\emptyset$.
 FH-bisimulation is defined formally: 
 \begin{definition}[Strong FH-bisimulation]\label{def-FH-bisim} ~\\
\noindent
\begin{minipage}{0.67\linewidth} 	Suppose 
   $A_1 = \mylangle J,\mathcal{S}_1, s_0,V_1,
   \mathcal{T}_1 \myrangle$ and $A_2 = \mylangle J,\mathcal{S}_2,t_0,V_2, \mathcal{T}_2 \myrangle$
   are open automata with identical holes of the same sort, with disjoint sets of variables.

@@

The definition of bisimulation for open automata allows us to derive the characterization and properties of a
bisimulation relation for open pNets. As pNets are composition
structures, it then makes sense to prove composition lemmas: we prove
that the composition of strongly bisimilar pNets are themselves
bisimilar.

@@

The initial state is defined as:
$InitState(\pNet) = \triangleleft {{s_i}_0}^{i\in L}  \triangleright$.
\end{definition}
To be precise, the state of each pLTS is entirely characterized by both the state of the automaton, and the value of its variables $V$. Consequently, the complete characterization of the state of a pNet should take into account the value of its variables $vars(P)$.

@@

Methodologies for reasoning about abstract semantics of open systems can be found in \cite{BaldanBB:2002, BaldanBB:07,Dubut:20}, authors introduce  behavioural equivalences for open systems from various symbolic approaches. Working in the setting of process calculi, some close relations  exist with the work  of the authors of \cite{BaldanBB:2002,BaldanBB:07}, where both approaches are based
on some kinds of labelled transition systems. The distinguishing feature of their approach is the transitions systems are labelled with logical formulae that provides  an abstract characterization of the structure that a hole must possess and of the actions it can perform in order to allow a transition to fire. Logical formulae are suitable formats that capture the general class of components that can act as the placeholders of the system during its evolution. In our approach we purposely leave the algebra of action terms undefined but the only operation we allow on action of holes is the comparison with other actions. Defining properly the interaction between a logical formulae in the action and the logics of the pNet composition seems very difficult.

@@

\noindent ($\Leftarrow$) We proceed by structural induction on the rules  building  the weak transition (as described in the original definition). The recurrence hypothesis being that the  the original definition implies the characterization (1),with the conditions stated at the bottom of the theorem. We consider the different rules:

@@

With the above lemma, we can use the same technique as for Theorem \ref{thm-equiv}  to prove that a weak FH-bisimulation is an equivalence. Indeed, we essentially use the same proof-scheme the main difference concerns  $\beta$ and  $\gamma$. Indeed, while the schema of the proof of transitivity was not directly applicable on the definition of weak bisimulation,  Lemma~\ref{lem-alternative-weak-bisim} provides a characterization of weak bisimulation similar to the definition of strong bisimulation, and thus the same proof scheme is directly applicable.
